Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-12 vertical-thermal-review.Rmd
wordsmithing VOC paragraphs
#102

<!--VOCs-->
Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017]. 
Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005]. 
In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017]. 
Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].

Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018]. 
However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021]. 
Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013]. 
This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient. 
Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.

>>>>

<!--VOCs-->
Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017]. 
In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017]. 
Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].

Less than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile is influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018]. 
However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021]. 
Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013]. 
This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient. 
Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
